Variational theory of non-relativistic quantum electrodynamics

### 2018-10-17 paper4pg.tex
some more corrections, I think we need some citations for the casimir related discussion

With the Hamiltonian described, we now move to develop a variational theory of the ground state. In the variational theorem, we choose an ansatz $|\Omega\rangle$ for the ground state of $H$. The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy. Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.  We choose as our ansatz
\begin{equation}
|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
\end{equation}
In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons (see Figure 1). Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions. Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$. Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$. We enforce that the matter and photon remain normalized by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),
\end{equation}
\end{widetext}
with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization conditions. Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  immediately yields:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x).
\end{align}
for the electron orbitals and energies. We see that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential. We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.

@@

Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz. In many quantum electrodynamical problems, this linear term. To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, two-photon spontaneous emission, and effective interactions betweeen distinct emitters. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Eq. (5) with states that simultaneously have virtual excitations of the matter and the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
\begin{figure}[t]
\includegraphics[width=5cm]{figure2new.pdf}
\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} Lowest two eigenvalues as a function of coupling for the multi-mode Rabi model describing the coupling of an emitter to a one-dimensional cavity. The variational results (blue) agree to within roughly a percent with the result derived by numerical diagonalization (orange) retaining up to four photons in the Hilbert space. The result from perturbation theory (green) disagrees strongly with the other methods of calculation for the first excited state due to a near-resonance between the bare transition energy and the bare lowest cavity mode.}
\label{fig:ansatz}
\end{figure}
We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by 
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{n=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)}
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs. (6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq. (8).
\begin{figure}[t]
\includegraphics[width=8.4cm]{figure3new.pdf}
\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule. (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization. These results show that our accuracy is robust to the number of levels in the system. }
\label{fig:ansatz}
\end{figure}

In what follows, we provide a proof-of-concept demonstration of the efficacy of the variational theory derived here. We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction. As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$. Working under the long-wavelength or dipole approximation, the Hamiltonian can then be written as:
\begin{equation}\label{eq:hamiltonian}
H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d).
\end{equation}
The fields can be expressed as a mode expansion, where for a cavity of length $L$, the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$. The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$. The matter system we describe as an $N_a$ site system. This could be considered as a simple model of a molecule within a tight-binding description. Thus we parameterize the general family of matter Hamiltonians as:
\begin{equation}\label{eq:matter_hamiltonian}
H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|) 
\end{equation}
The momentum operator, we write as 
\begin{equation}\label{eq:momentum_operator}
p = \frac{-i\hbar}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
\end{equation}
where $R$ is a constant with units of length representing roughly the difference in positions between sites. This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced. In other words: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.  Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system. However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot. It thus places a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be. The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.

The detailed calculations of the energies of states via the formalism introduced here will be shown elsewhere. Here, we state the main results. Using an essentially one-dimensional version of Eq. (7), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations. In the absence of correlations, we found that the ground state energy satisfies
\begin{equation}
E = E_g + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
\end{equation}
where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$. This is a remarkable result, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system. In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart. This is simply because the electromagnetic mode structure of two nearby plates is very different from that of two infinitely separated plates. This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies. This logic can be applied to any arrangement of macroscopic polarizable objects. What is remarkable about the result of Eq. (13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.

In the presence of correlations  we must add to the energy a contribution of the form of Eq. (8), specialized to the case of an emitter in a one-dimensional cavity. We can actually apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Eqs. (6) and (7). It is worthwhile to note that for this one-dimensional case, as we show elsewhere, we arrive at semi-analytic results for all variational energies that we calculate.

@@

This result very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime. Moreover, our results demonstrate a successful non-perturbative theory of the Lamb shift and consequently Casimir-Polder forces. In particular, it is long known that energy levels of emitters can shift as a result of virtual photon emission and re-absorption. These energy shifts, called Lamb shifts, depend on the particular position of the emitter in the photonic structure it is embedded in. These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces. Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq. (8) using bare atomic and photonic properties. Thus, our calculations, which find the energy shifts non-perturbatively, lead to a consequently non-perturbative theory of Casimir-Polder forces.

>>>>

With the Hamiltonian described, we now move to develop a variational theory of the ground state. In the variational theorem, we choose an ansatz $|\Omega\rangle$ for the ground state of $H$. The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy. Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.  We choose as our ansatz
\begin{equation}
|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
\end{equation}
In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons (see Figure 1). Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions. Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$. Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$. We enforce that the matter and photon remain normalized by constructing the Lagrange function:
\begin{align}
&\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle \\
&- \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),\nonumber
\end{align}
with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization conditions. Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  immediately yields:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x).
\end{align}
for the electron orbitals and energies. We see that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential. We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.

@@

Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz. To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, two-photon spontaneous emission, and effective interactions betweeen distinct emitters \Jadd{(Nick: here we need citations)}. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Eq. (5) with states that simultaneously have virtual excitations of the matter and the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
\begin{figure}[t]
\includegraphics[width=5cm]{figure2new.pdf}
\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} Lowest two eigenvalues as a function of coupling for the multi-mode Rabi model describing the coupling of an emitter to a one-dimensional cavity. The variational results (blue) agree to within roughly a percent with the result derived by numerical diagonalization (orange) retaining up to four photons in the Hilbert space. The result from perturbation theory (green) disagrees strongly with the other methods of calculation for the first excited state due to a near-resonance between the bare transition energy and the bare lowest cavity mode.}
\label{fig:ansatz}
\end{figure}
We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by 
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{n=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)}
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs. (6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq. (8).
\begin{figure}[t]
\includegraphics[width=8.4cm]{figure3new.pdf}
\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule. (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization. These results show that our accuracy is robust to the number of levels in the system. }
\label{fig:ansatz}
\end{figure}

In what follows, we provide a proof-of-concept demonstration of the efficiency of the variational theory derived here. We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction. As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$. Working under the long-wavelength or dipole approximation, the Hamiltonian can then be written as:
\begin{equation}\label{eq:hamiltonian}
H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d).
\end{equation}
The fields can be expressed as a mode expansion, where for a cavity of length $L$, the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$. The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$. The matter system we describe as an $N_a$ site system. This could be considered as a simple model of a molecule within a tight-binding description. Thus we parameterize the general family of matter Hamiltonians as:
\begin{equation}\label{eq:matter_hamiltonian}
H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|) 
\end{equation}
The momentum operator, we write as 
\begin{equation}\label{eq:momentum_operator}
p = \frac{-i\hbar}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
\end{equation}
where $R$ is a constant with units of length representing roughly the difference in positions between sites. This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced. In other words: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.  Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system. However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot. It thus places a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be. The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.

The detailed calculations of the energies of states via the formalism introduced here will be shown elsewhere. Here, we state the main results. Using an essentially one-dimensional version of Eq. (7), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations. In the absence of correlations, we found that the ground state energy satisfies
\begin{equation}
E = E_g + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
\end{equation}
where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$. This is a remarkable result, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system. In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart. This is simply because the electromagnetic mode structure of two nearby plates is very different from that of two infinitely separated plates. This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies. This logic can be applied to any arrangement of macroscopic polarizable objects. What is remarkable about the result of Eq. (13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object. 
\Jadd{(Nick, the upper section would benefit from 1 or 2 citations)}
In the presence of correlations  we must add to the energy a contribution of the form of Eq. (8), specialized to the case of an emitter in a one-dimensional cavity. We can actually apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Eqs. (6) and (7). It is worthwhile to note that for this one-dimensional case, as we show elsewhere, we arrive at semi-analytic results for all variational energies that we calculate.

@@

This result very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime. Moreover, our results demonstrate a successful non-perturbative theory of the Lamb shift and consequently Casimir-Polder forces. In particular, it is long known that energy levels of emitters can shift as a result of virtual photon emission and re-absorption. These energy shifts, called Lamb shifts, depend on the particular position of the emitter in the photonic structure it is embedded in. These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces. Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory\Jadd{(Nick, cite)}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq. (8) using bare atomic and photonic properties. Thus, our calculations, which find the energy shifts non-perturbatively, lead to a consequently non-perturbative theory of Casimir-Polder forces \Jadd{(Nick, cite)}.
